Visualizing probability distributions across bivariate cyclic temporal granularities

### 2021-05-01 main.Rmd
gram and spell check

Temporal data are available at various resolutions depending on the context. Social and economic data are often collected and reported at coarse temporal scales such as monthly, quarterly or annually. With recent advancement in technology, more and more data are recorded at much finer temporal scales. Energy consumption may be collected every half an hour, energy supply may be collected every minute, and web search data might be recorded every second. As the frequency of data increases, the number of questions about the periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales such as days, months or quarters. This approach requires deconstructing time in various possible ways called time granularities [@aigner2011visualization].

@@

Calendar-based graphics [@wang2020calendar] are useful in visualizing patterns in the weekly and monthly structure, and are helpful when checking for the effects of weekends or special days. Any temporal data at sub-daily resolution can also be displayed using this type of faceting [@Wickham2009pk] with days of the week, month of the year, or another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week or month-of-year deconstructions. There can be many different time deconstructions, based on the calendar or on categorizations of time granularities.

@@

For example, day is periodic with respect to week with $R=1$ and $P=7$, while (if we ignore leap years) day is periodic with respect to month with $R=12$ and $P=365$ as any month would consist of same number of days across years.
Since the idea of period involves a pair of granularities, we say that the pair $(day, week)$ has period 7, while the pair $(day, month)$ has a period 365 (ignoring leap years).

@@

\autoref{fig:circular-dow} illustrates some linear and cyclical granularities. Cyclical granularities are constructed be cutting the linear granularity into pieces, and stacking them to match the cycles (as shown in b). $B, G, H$ (day, week, fortnight, respectively) are linear granularities. The circular granularity $C_{B, G}$ (day-of-week) is constructed from $B$ and $G$, while  circular granularity $C_{B, H}$ (day-of-fortnight) is constructed from $B$ and $H$. These overlapping cyclical granularities share elements from the linear granularity. Each of $C_{B , G}$ and $C_{B , H}$ consist of repeated patterns $\{0, 1, \dots, 6\}$ and $\{0, 1, \dots, 13\}$ with $P=7$ and $P=14$ respectively.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lll}
\toprule
Circular granularity & Expression & Period \\
\midrule
minute-of-hour                               &
  $C_1 = z \mod 60$                     &
  $P_1 = \phantom{99}60$ \\
minute-of-day                                &
  $C_j = z \mod 60*24$                  &
  $P_2= `r 60*24`$\\
hour-of-day                                  &
  $C_3 = \lfloor z/60\rfloor\mod 24$    &
  $P_3 = \phantom{99}24$ \\
hour-of-week                                 &
  $C_4 = \lfloor z/60\rfloor\mod 24*7$  &
  $P_4= \phantom{9}`r 24*7`$\\
day-of-week                                  &
  $C_5 = \lfloor z/24*60\rfloor \mod 7$ &
  $P_5= \phantom{999}7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of circular granularities with bottom granularity minutes. Circular granularity $C_i$ relates two linear granularities one of which groups periodically into the other with regular mapping and period $P_i$. Circular granularities can be expressed using modular arithmetic due to their regular mapping. }
\label{tab:definitions}
\end{table}

@@

(ref:aperiodic-example) Quasi-circular and aperiodic cyclic granularities illustrated through (a) linear and (b) stacked displays of time. The linear display shows granularities days ($B$), weeks ($G$), semester days ($B'$), and stages of a semester ($M$) distributed over linear time. Granules of $B'$ is only defined for days when the semester is running. Here a semester lasts for 18 weeks and 2 days, each consisting of 5 stages. It starts with an orientation stage, followed by a first stage of in-session period (6 weeks), a break (1 week), a second stage of in-session period (7 weeks), a 1-week study break stage, before final exams stage which continue for 16 days. This distribution of semester days within a semester remains the same for every semesters and hence $Q_{B', M}$ with $P$ = 128 days will be a quasi-circular granularity with repeating patterns. $A_{B, M}$ will be an aperiodic cyclic granularity since 
the placement of the semester within an year varies across years and there is no definite number of days after which the semester begins every year.

The linear display shows granularities days ($B$), weeks ($G$), , and stages of a semester ($M$) distributed over linear time.

For example, consider the school semester shown in \autoref{fig:aperiodic-example}. Let the linear granularities $M_1$ and $M_2$ denote the teaching and non-teaching stages of the semester respectively. Both $M_1$, $M_2$ and $M = M_{1}\bigcup M_{2}$ denoting the "stages" of the semester are aperiodic with respect to days ($B$) or weeks ($G$). Hence $A_{B, M}$ denoting day-of-the-stage would be an aperiodic cyclic granularity, because the placement of the semester within an year would vary across years. Here, $Q_{B', M}$ denoting semester-day-of-the-stage would be a quasi-circular granularity 
since the distribution of semester days within a semester is assumed to remain constant over years. Here semester-day is denoted by "sem day" ($B'$) and its granules are only defined for the span of the semesters.

@@

Like most calendars, the Mayan calendar used the day as the basic unit of time [@Reingold2001-kf]. The structuring of larger units, weeks, months, years and cycle of years, though, varies substantially between calendars. For example,  the French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds", the duration of which is 0.864 common seconds. Nevertheless, for any calendar a hierarchy table can be defined. Note that it is not always possible to organize an aperiodic linear granularity in a hierarchy table. Hence, we assume that the hierarchy table consists of periodic linear granularities only, and that the cyclic granularity $C_{G(\ell),G(m)}$ is either circular or quasi-circular.

@@

Effective exploration and visualization benefits from well-organized data structures. @wang2020tsibble introduced the tidy "tsibble" data structure to support exploration and modeling of temporal data. This forms the basis of the structure for cyclic granularities. A tsibble comprises an index, optional key(s), and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble.

@@

An example of a harmony is where $C_i$ and $C_j$ denote day-of-week and month-of-year respectively. So $C_i$ will have 7 levels while $C_j$ will have 12 levels, giving $12\times7=84$ sets $S_{k\ell}$. All of these are non-empty because every day-of-week can occur in every month. Hence, the pair (day-of-week, month-of-year) is a harmony.

@@

The purpose is to visualize the distribution of the continuous variable ($v$) conditional on the values of two granularities , $C_i$ and $C_j$. Since $C_i$ and $C_j$ are factors or categorical variables, data subsets corresponding to each combination of their levels form a subgroup and the visualization amounts to having displays of distributions for different subgroups. The response variable $(v)$ is plotted on the y-axis and the levels of $C_i (C_j)$ on the x-axis, conditional on the levels of $C_j (C_i)$. This means, carrying out the same plot corresponding to each level of the conditioning variable. This is consistent with the widely used grammar of graphics
which is a framework to construct statistical graphics by relating the data space to the graphic space [@Wilkinson1999-nk, @Wickham2009pk].

@@

There are several ways to summarize a distribution like estimating empirical distribution or density of the variable's distribution, computing few number summaries or quantiles. This data estimation or summarization could be potentially misleading if they are performed on rarely occurring categories (Section&nbsp;\ref{sec:near-clashes}). Even when there are no rarely occurring events, the number of observations may vary greatly within or across each facet, due to missing observations or uneven locations of events in the time domain. In such cases, data summarization should be used with caution as sample sizes will directly affect the accuracy of the estimated quantities being displayed.

@@

The basic plot choice for our data structure is one that can display distributions. For displaying the distribution of continuous univariate variable, many options are available. Displays based on descriptive statistics include box plots [@Tukey1977-jx] and its variation such as notched box plots [@McGill1978-hg] or other variations as mentioned in @Wickham2012-vr. They also include line or area quantile plots which can display any quantiles and not only quartiles like in a boxplot. Plots based on kernel density estimates include violin plots [@Hintze1998-zi], summary plot [@Potter2010-qc], ridge line plots [@R-ggridges], and highest density region (HDR) plots [@Hyndman1996-ft]. The less commonly used Letter-Value plots [@Hofmann2017-sg] is a midway between boxplots and density plots. Letter values are order statistics with specific depths, for example, the median ($M$) is a letter value which divides the data set into halves. Each of the next letter values splits the remaining parts into two separate regions so that the fourths ($F$), eighths ($E$), sixteenths ($D$) etc. are obtained. They are useful for displaying the distributions beyond the quartiles specially for large data,
where boxplots mislabel data points as outliers.
One of the best approaches in exploratory data analysis is to draw a variety of plots to reveal information while keeping in mind the drawbacks and benefits of each of the plot choices. For example, boxplots obscure multimodality, and interpretation of density estimates and histograms may change depending on the bandwidth and binwidths respectively. In R package `gravitas` [@R-gravitas], boxplots, violin, ridge, letter-value, line and area quantile plots are implemented, but it is potentially possible to use any plots which can display the distribution of the data.

@@

(ref:allFig) Distribution of energy consumption displayed through letter value plots: plot a across month-of-year faceted by quarter-of-year; plot b across weekday/weekend faceted by quarter-of-year; plot c across quarter-of-year faceted by weekday/weekend. Plot a shows a clash since there are empty combinations. Plots b and c show harmonies since each quarter includes both weekdays and weekends. It can be seen in plot b that for every quarter (mapped to facet), weekend and weekday consumption are fairly similar except for the second quarter where the letter values below D and E behave differently. This is probably because of lower temperatures in the second quarter compared to the first quarter (summer in Australia). Plot c switches the granularities mapped to the x-axis and the facets and helps to compare quarters within weekdays and weekends. 
<!-- For example, for weekdays the interquartile range of consumption reduces over the year, whereas this pattern is not true for weekends. -->

@@

When $C_i$ is mapped to the $x$ position and $C_j$ to facets, then the $A_k$ levels are juxtaposed and each $B_\ell$ represent a group/facet. Gestalt theory suggests that when items are placed in close proximity, people assume that they are in the same group because they are close to one another and apart from other groups. Hence, in this case the $A_k$'s are compared against each other within each group. With the mapping of $C_i$ and $C_j$ reversed, the emphasis will shift to comparing $B_\ell$ levels rather than $A_k$ levels. For example, \autoref{fig:allFig}b shows the letter value plot across weekday/weekend partitioned by quarters of the year and \autoref{fig:allFig}c shows the same two cyclic granularities with their mapping reversed. \autoref{fig:allFig}b helps us to compare weekday and weekend within each quarter and \autoref{fig:allFig}c helps to compare quarters within weekend and weekday.

@@

To identify the available cyclic time granularities, consider the conventional time deconstructions for a Gregorian calendar that can be formed from the 30-minute time index: half-hour, hour, day, week, month, quarter, half-year, year. In this example, we will consider the granularities hour, day, week and month giving six cyclic granularities "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. To these we add day-type ("wknd_wday") to capture weekend and weekday behavior. Now that we have a list of cyclic granularities to look at, we can compute them using the results in Section&nbsp;\ref{sec:cyclic-calendar}.

@@

(ref:bothcust) Energy consumption of a single customer shown with different distribution displays, and granularity arrangements. Two granularities are used: 1. hour of the day; and 2. weekday/weekend. Plot a shows granularity 2 faceted by granularity 1; plots b and c shows the reverse mapping. Plot a allows comparison of usage by workday within each hour of the day using side-by-side boxplots, showing that on work days there is more consumption early in the day. Plots b and c examine the temporal trend of consumption over the course of a day, separately for the type of day. Plot b uses an area quantile to emphasize the time component; e.g., median consumption shows prolonged high usage in the morning on weekdays. Plot c uses a violin plot emphasize subtler distributional differences across hours, showing that morning use on weekdays is bimodal, and on some work days there is low usage possibly indicating the person is working from home or having a late start.

@@

The same data is shown using violin plots instead of quantile plots in \autoref{fig:bothcust}c. There is bimodality in the early hours of the day for weekdays and weekends. If we visualize the same data with reverse mapping of the cyclic granularities (\autoref{fig:bothcust}a), then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. Then it can be seen that median energy consumption for the early morning hours is higher for weekdays than weekends. Also, outliers are more prominent in the latter hours of the day. All of these indicate that looking at different distribution geometry or changing the mapping can shed light on different aspects of energy behavior for the same sample.

@@

Exploratory data analysis involve many iterations of finding and summarizing patterns. With temporal data available at ever finer scales, exploring periodicity can become overwhelming with so many possible granularities to explore. This work provides tools to classify and compute possible cyclic granularities from an ordered (usually temporal) index. We also provide a framework to systematically explore the distribution of a univariate variable conditional on two cyclic time granularities using visualizations based on the synergy and levels of the cyclic granularities.

@@

We propose producing plots based on pairs of cyclic granularities that form harmonies rather than clashes or near-clashes. A future direction of work could be to further refine the selection of appropriate pairs of granularities by identifying those for which the differences between the displayed distributions is greatest, and rating these selected harmony pairs in order of importance for exploration.

>>>>

Temporal data are available at various resolutions depending on the context. Social and economic data are often collected and reported at coarse temporal scales such as monthly, quarterly or annually. With recent advancements in technology, more and more data are recorded at much finer temporal scales. Energy consumption may be collected every half an hour, energy supply may be collected every minute, and web search data might be recorded every second. As the frequency of data increases, the number of questions about the periodicity of the observed variable also increases. For example, data collected at an hourly scale can be analyzed using coarser temporal scales such as days, months or quarters. This approach requires deconstructing time in various possible ways called time granularities [@aigner2011visualization].

@@

Calendar-based graphics [@wang2020calendar] are useful in visualizing patterns in the weekly and monthly structure and are helpful when checking for the effects of weekends or special days. Any temporal data at sub-daily resolution can also be displayed using this type of faceting [@Wickham2009pk] with days of the week, month of the year, or another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week or month-of-year deconstructions. There can be many different time deconstructions, based on the calendar or on categorizations of time granularities.

@@

For example, day is periodic with respect to week with $R=1$ and $P=7$, while (if we ignore leap years) day is periodic with respect to month with $R=12$ and $P=365$ as any month would consist of the same number of days across years.
Since the idea of period involves a pair of granularities, we say that the pair $(day, week)$ has period 7, while the pair $(day, month)$ has a period 365 (ignoring leap years).

@@

\autoref{fig:circular-dow} illustrates some linear and cyclical granularities. Cyclical granularities are constructed by cutting the linear granularity into pieces, and stacking them to match the cycles (as shown in b). $B, G, H$ (day, week, fortnight, respectively) are linear granularities. The circular granularity $C_{B, G}$ (day-of-week) is constructed from $B$ and $G$, while  circular granularity $C_{B, H}$ (day-of-fortnight) is constructed from $B$ and $H$. These overlapping cyclical granularities share elements from the linear granularity. Each of $C_{B , G}$ and $C_{B , H}$ consist of repeated patterns $\{0, 1, \dots, 6\}$ and $\{0, 1, \dots, 13\}$ with $P=7$ and $P=14$ respectively.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lll}
\toprule
Circular granularity & Expression & Period \\
\midrule
minute-of-hour                               &
  $C_1 = z \mod 60$                     &
  $P_1 = \phantom{99}60$ \\
minute-of-day                                &
  $C_j = z \mod 60*24$                  &
  $P_2= `r 60*24`$\\
hour-of-day                                  &
  $C_3 = \lfloor z/60\rfloor\mod 24$    &
  $P_3 = \phantom{99}24$ \\
hour-of-week                                 &
  $C_4 = \lfloor z/60\rfloor\mod 24*7$  &
  $P_4= \phantom{9}`r 24*7`$\\
day-of-week                                  &
  $C_5 = \lfloor z/24*60\rfloor \mod 7$ &
  $P_5= \phantom{999}7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of circular granularities with bottom granularity minutes. Circular granularity $C_i$ relates two linear granularities one of which groups periodically into the other with regular mapping and period $P_i$. Circular granularities can be expressed using modular arithmetic due to their regular mapping.}
\label{tab:definitions}
\end{table}

@@

(ref:aperiodic-example) Quasi-circular and aperiodic cyclic granularities illustrated through (a) linear and (b) stacked displays of time. The linear display shows granularities days ($B$), weeks ($G$), semester days ($B'$), and stages of a semester ($M$) distributed over linear time. Granules of $B'$ is only defined for days when the semester is running. Here a semester lasts for 18 weeks and 2 days, each consisting of 5 stages. It starts with an orientation stage, followed by a first stage of the in-session period (6 weeks), a break (1 week), a second stage of the in-session period (7 weeks), a 1-week study break stage, before final exams stage which continue for 16 days. This distribution of semester days within a semester remains the same for every semester and hence $Q_{B', M}$ with $P$ = 128 days will be a quasi-circular granularity with repeating patterns. $A_{B, M}$ will be an aperiodic cyclic granularity since 
the placement of the semester within a year varies across years and there is no definite number of days after which the semester begins every year.

For example, consider the school semester shown in \autoref{fig:aperiodic-example}. Let the linear granularities $M_1$ and $M_2$ denote the teaching and non-teaching stages of the semester respectively. Both $M_1$, $M_2$ and $M = M_{1}\bigcup M_{2}$ denoting the "stages" of the semester are aperiodic with respect to days ($B$) or weeks ($G$). Hence $A_{B, M}$ denoting day-of-the-stage would be an aperiodic cyclic granularity because the placement of the semester within a year would vary across years. Here, $Q_{B', M}$ denoting semester-day-of-the-stage would be a quasi-circular granularity
since the distribution of semester days within a semester is assumed to remain constant over years. Here semester-day is denoted by "sem day" ($B'$) and its granules are only defined for the span of the semesters.

@@

Like most calendars, the Mayan calendar used the day as the basic unit of time [@Reingold2001-kf]. The structuring of larger units, weeks, months, years and cycle of years, though, varies substantially between calendars. For example,  the French revolutionary calendar divided each day into 10 "hours", each "hour" into 100 "minutes" and each "minute" into 100 "seconds", the duration of which is 0.864 common seconds. Nevertheless, for any calendar, a hierarchy table can be defined. Note that it is not always possible to organize an aperiodic linear granularity in a hierarchy table. Hence, we assume that the hierarchy table consists of periodic linear granularities only, and that the cyclic granularity $C_{G(\ell),G(m)}$ is either circular or quasi-circular.

@@

Effective exploration and visualization benefit from well-organized data structures. @wang2020tsibble introduced the tidy "tsibble" data structure to support exploration and modeling of temporal data. This forms the basis of the structure for cyclic granularities. A tsibble comprises an index, optional key(s), and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble.

@@

An example of harmony is where $C_i$ and $C_j$ denote day-of-week and month-of-year respectively. So $C_i$ will have 7 levels while $C_j$ will have 12 levels, giving $12\times7=84$ sets $S_{k\ell}$. All of these are non-empty because every day-of-week can occur in every month. Hence, the pair (day-of-week, month-of-year) is a harmony.

@@

The purpose is to visualize the distribution of the continuous variable ($v$) conditional on the values of two granularities, $C_i$ and $C_j$. Since $C_i$ and $C_j$ are factors or categorical variables, data subsets corresponding to each combination of their levels form a subgroup and the visualization amounts to having displays of distributions for different subgroups. The response variable $(v)$ is plotted on the y-axis and the levels of $C_i (C_j)$ on the x-axis, conditional on the levels of $C_j (C_i)$. This means, carrying out the same plot corresponding to each level of the conditioning variable. This is consistent with the widely used grammar of graphics
which is a framework to construct statistical graphics by relating the data space to the graphic space [@Wilkinson1999-nk, @Wickham2009pk].

@@

There are several ways to summarize a distribution like estimating empirical distribution or density of the variable's distribution, computing a few number summaries or quantiles. This data estimation or summarization could be potentially misleading if they are performed on rarely occurring categories (Section&nbsp;\ref{sec:near-clashes}). Even when there are no rarely occurring events, the number of observations may vary greatly within or across each facet, due to missing observations or uneven locations of events in the time domain. In such cases, data summarization should be used with caution as sample sizes will directly affect the accuracy of the estimated quantities being displayed.

@@

The basic plot choice for our data structure is one that can display distributions. For displaying the distribution of a continuous univariate variable, many options are available. Displays based on descriptive statistics include box plots [@Tukey1977-jx] and its variants such as notched box plots [@McGill1978-hg] or other variations as mentioned in @Wickham2012-vr. They also include line or area quantile plots which can display any quantiles and not only quartiles like in a boxplot. Plots based on kernel density estimates include violin plots [@Hintze1998-zi], summary plot [@Potter2010-qc], ridge line plots [@R-ggridges], and highest density region (HDR) plots [@Hyndman1996-ft]. The less commonly used Letter-Value plots [@Hofmann2017-sg] is midway between boxplots and density plots. Letter values are order statistics with specific depths, for example, the median ($M$) is a letter value that divides the data set into halves. Each of the next letter values splits the remaining parts into two separate regions so that the fourths ($F$), eighths ($E$), sixteenths ($D$), etc. are obtained. They are useful for displaying the distributions beyond the quartiles especially for large data,
where boxplots mislabel data points as outliers.
One of the best approaches in exploratory data analysis is to draw a variety of plots to reveal information while keeping in mind the drawbacks and benefits of each of the plot choices. For example, boxplots obscure multimodality, and interpretation of density estimates and histograms may change depending on the bandwidth and binwidths respectively. In R package `gravitas` [@R-gravitas], boxplots, violin, ridge, letter-value, line and area quantile plots are implemented, but it is potentially possible to use any plots which can display the distribution of the data.

@@

(ref:allFig) Distribution of energy consumption displayed through letter value plots: plot a across month-of-year faceted by quarter-of-year; plot b across weekday/weekend faceted by quarter-of-year; plot c across quarter-of-year faceted by weekday/weekend. Plot a shows a clash since there are empty combinations. Plots b and c show harmonies since each quarter include both weekdays and weekends. It can be seen in plot b that for every quarter (mapped to facet), weekend and weekday consumption are fairly similar except for the second quarter where the letter values below D and E behave differently. This is probably because of lower temperatures in the second quarter compared to the first quarter (summer in Australia). Plot c switches the granularities mapped to the x-axis and the facets and helps to compare quarters within weekdays and weekends. 
<!-- For example, for weekdays the interquartile range of consumption reduces over the year, whereas this pattern is not true for weekends. -->

@@

When $C_i$ is mapped to the $x$ position and $C_j$ to facets, then the $A_k$ levels are juxtaposed and each $B_\ell$ represents a group/facet. Gestalt theory suggests that when items are placed in close proximity, people assume that they are in the same group because they are close to one another and apart from other groups. Hence, in this case, the $A_k$'s are compared against each other within each group. With the mapping of $C_i$ and $C_j$ reversed, the emphasis will shift to comparing $B_\ell$ levels rather than $A_k$ levels. For example, \autoref{fig:allFig}b shows the letter value plot across weekday/weekend partitioned by quarters of the year and \autoref{fig:allFig}c shows the same two cyclic granularities with their mapping reversed. \autoref{fig:allFig}b helps us to compare weekday and weekend within each quarter and \autoref{fig:allFig}c helps to compare quarters within weekend and weekday.

@@

To identify the available cyclic time granularities, consider the conventional time deconstructions for a Gregorian calendar that can be formed from the 30-minute time index: half-hour, hour, day, week, month, quarter, half-year, year. In this example, we will consider the granularities hour, day, week and month giving six cyclic granularities "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. To these, we add day-type ("wknd_wday") to capture weekend and weekday behavior. Now that we have a list of cyclic granularities to look at, we can compute them using the results in Section&nbsp;\ref{sec:cyclic-calendar}.

@@

(ref:bothcust) Energy consumption of a single customer shown with different distribution displays, and granularity arrangements. Two granularities are used: 1. hour of the day; and 2. weekday/weekend. Plot a shows granularity 2 faceted by granularity 1; plots b and c shows the reverse mapping. Plot a allows comparison of usage by workday within each hour of the day using side-by-side boxplots, showing that on work days there is more consumption early in the day. Plots b and c examine the temporal trend of consumption over the course of a day, separately for the type of day. Plot b uses an area quantile to emphasize the time component; e.g., median consumption shows prolonged high usage in the morning on weekdays. Plot c uses a violin plot emphasizes subtler distributional differences across hours, showing that morning use on weekdays is bimodal, and on some work days there is low usage possibly indicating the person is working from home or having a late start.

@@

The same data is shown using violin plots instead of quantile plots in \autoref{fig:bothcust}c. There is bimodality in the early hours of the day for weekdays and weekends. If we visualize the same data with reverse mapping of the cyclic granularities (\autoref{fig:bothcust}a), then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. Then it can be seen that median energy consumption for the early morning hours is higher for weekdays than weekends. Also, outliers are more prominent in the later hours of the day. All of these indicate that looking at different distribution geometry or changing the mapping can shed light on different aspects of energy behavior for the same sample.

@@

Exploratory data analysis involves many iterations of finding and summarizing patterns. With temporal data available at ever finer scales, exploring periodicity can become overwhelming with so many possible granularities to explore. This work provides tools to classify and compute possible cyclic granularities from an ordered (usually temporal) index. We also provide a framework to systematically explore the distribution of a univariate variable conditional on two cyclic time granularities using visualizations based on the synergy and levels of the cyclic granularities.

@@

We propose producing plots based on pairs of cyclic granularities that form harmonies rather than clashes or near-clashes. A future direction of work could be to further refine the selection of appropriate pairs of granularities by identifying those for which the differences between the displayed distributions is greatest and rating these selected harmony pairs in order of importance for exploration.
